
<!--
## 4. Simulate data/ Add missing taxa
Finally, if data can be added to An alternative to generate a dated tree from a set of taxa is to take the available
information and simulate into it the missing data.
We will take the median and sdm summary chronograms to date the Synthetic tree of Life:
```{r LTTplot-sims, echo = F, results = 'asis', fig.cap= ""}
#nothing yet
```
-->

\newpage

## 4. Example with subspecies tree
As an example, we're gonna date the subspecies tree of the group using all approaches
for generating new data.
```{r echo = FALSE}
# txtA <- toupper("Something is missing.")
# if(!inherits(tax_allcal_datedotol, "list")){
#   txtA <- paste(taxon, "and no tree could be constructed.")
# } else {
#     if(is.null(tax_allcal_datedotol$phy)){
#       txtA <- paste(taxon, "and they do not agree, so no tree could be constructed.")
#     } else {
#       if(all(is.na(tax_allcal_datedotol$phy$edge.length))){
#         txtA <- paste(taxon, "and a tree was constructed, but all branch lengths are NA.")
#       } else {
#           txtA <- paste(taxon, "and a tree with", ape::Ntip(tax_allcal_datedotol$phy), "tips,",
#                         round(tax_allcal_datedotol$phy$Nnode/ape::Ntip(tax_allcal_datedotol$phy)*100),
#                         "% resolved nodes and a MRCA of",
#                         round(max(ape::branching.times(tax_allcal_datedotol$phy))),"was constructed.")
#       }
#     }
# }
```

Now, let's say you like the Open Tree of Life Taxonomy and you want to stick to
that tree. Dates from available studies were tested over the Open Tree of Life
Synthetic tree of `r txtA`
We also tried  each source chronogram independently, with the Dated OToL and with
each other, as a form of cross validation in `r table_crossval`. This is not working
perfectly yet, but we are developping new ways to use all calibrations efficiently.

\newpage


<!--
\newpage

```{r echo = FALSE}
table_cap <- "Was it successful to use each source chronogram independently as calibration (CalibN) against the Dated Open Tree of Life (dOToL) and each other (ChronoN)?"
table2 <- cbind(tax_eachcal_datedotol, do.call(cbind, tax_crossval))
dimnames(table2) <- list(paste0("Calibrations", seq(tax_crossval)), c("dOToL", paste0("Chrono", seq(tax_crossval))))
if(ncol(table2) < 7) fs <- 10
if(ncol(table2) == 7 | ncol(table2) == 8) fs <- 9
if(ncol(table2) >= 9) fs <- 7
if(ncol(table2) > 10) {
  table2ori <- table2
  table2[table2 == TRUE] <- "yes" # "&#x2611;"  # :tick:
  table2[table2 == "FALSE"] <- "no"  # "&#x2612;"
  dimnames(table2) <- list(paste0("Calib", seq(tax_crossval)),
                           c("dOToL", paste0("Chr", seq(tax_crossval))))
}
tt <- knitr::kable(table2, caption = table_cap, row.names = TRUE, format = "latex", booktabs = TRUE)
kableExtra::kable_styling(kable_input = tt, position = "left", font_size = fs)
```
-->
```{r LTTplot-phyloall, echo = F, results = 'asis', fig.cap= ""}
figcap_lttplot_phyloall <- paste0("Lineage through time (LTT) plots of source chronograms
  available in database for species in the ", taxon, ". Numbers correspond to original
  studies in ", table_chronograms, ". Arrows indicate maximum age of each chronogram.")
lttplot_phyloall <- paste0("\n\\begin{figure}[h!]\n",
    "\t\\includegraphics{plots/", taxon, "_LTTplot_phyloall.pdf}\n",
    "\t\\caption{", figcap_lttplot_phyloall, "}\n",
    "\t\\label{fig:lttplot_phyloall}\n",
    "\\end{figure}")
cat(lttplot_phyloall)
```

\newpage

```{r LTTplot-cluster-median, echo = F, results = 'asis', fig.cap= ""}
figcap_lttplot <- paste0("Lineage Through Time plots of ", taxon, " median summary
chronograms obtained with different clustering algorithms. Not all algorithms worked
with this summary matrix and we are only showing here the ones that worked.
Chronograms obtained from the SDM summary matrix are very similar to the ones from
the median summary matrix with all clustering algorithms (Appendix Fig. \\ref{fig:lttplot_cluster_sdm}).")

lttplot_cluster <- paste0("\n\\begin{figure}[h!]\n",
    "\t\\includegraphics{plots/", taxon, "_lttplot_cluster_median.pdf}\n",
    "\t\\caption{", figcap_lttplot, "}\n",
    "\t\\label{fig:lttplot_cluster_median}\n",
    "\\end{figure}")
cat(lttplot_cluster)
```

\newpage

```{r LTTplot-summ-chronograms, echo = F, results = 'asis', fig.cap= ""}
figcap_lttplot <- paste(taxon, "lineage through time (LTT) plots of summary chronograms
    obtained by calibrating a consensus tree tropology with distance data
    from median (green) and SDM (blue) summary matrices and then adjusting branch
    lengths with BLADJ. Source chronograms are shown in gray for comparison.")

lttplot_summchrono <- paste0("\n\\begin{figure}[h!]\n",
    "\t\\includegraphics{plots/", taxon, "_LTTplot_summary_chronograms.pdf}\n",
    "\t\\caption{", figcap_lttplot, "}\n",
    "\t\\label{fig:lttplot_summchrono}\n",
    "\\end{figure}")
cat(lttplot_summchrono)
```

\newpage

```{r LTTplot-bladj, echo = F, results = 'asis', fig.cap= ""}
figcap_lttplot <- paste(taxon, "lineage through time (LTT) plots from
    source chronograms used as secondary calibrations (gray), source chronograms
    used as topology (purple) and chronograms resulting from calibrating the latter
    with the former, using BLADJ (green).")

lttplot <- paste0("\n\\begin{figure}[h!]\n",
    "\t\\includegraphics{plots/", taxon, "_LTTplot_crossval_bladj.pdf}\n",
    "\t\\caption{", figcap_lttplot, "}\n",
    "\t\\label{fig:lttplot_crossval_bladj}\n",
    "\\end{figure}")
cat(lttplot)
```

\newpage

```{r LTTplot-pathd8-exp1, echo = F, results = 'asis', fig.cap= ""}
figcap_lttplot <- paste(taxon, "lineage through time (LTT) plots from
    source chronograms used as secondary calibrations (gray), source chronograms
    used as topology (purple) and chronograms resulting from calibrating the latter
    with the former, using PATHd8 (green).")

lttplot <- paste0("\n\\begin{figure}[h!]\n",
    "\t\\includegraphics{plots/", taxon, "_LTTplot_crossval_pathd8_exp1.pdf}\n",
    "\t\\caption{", figcap_lttplot, "}\n",
    "\t\\label{fig:lttplot_crossval_pathd8_exp1}\n",
    "\\end{figure}")
cat(lttplot)
```

\break
\newpage













# Appendix



```{r myfile-1-plot, echo = F, results = 'asis', fig.cap= ""}
# knitr::include_graphics(paste0("docs/datedotol_", taxon, ".pdf"))
figcap <- paste(taxon, "Species Dated Open Tree of Life Induced Subtree. This chronogram was obtained with `get_dated_otol_induced_subtree()` function.")

if(!inherits(tax_datedotol, "phylo")){
  plot1 <- paste0("Dated induced subtree could not be obtained for the ", taxon, ".\n")
} else {
  plot1 <- paste0("\n![", figcap, "](plots/", taxon, "_datedotol.pdf)\n")
}
cat(plot1)
```


```{r eval = TRUE, echo = F, results = 'asis'}
figcap_lttplot_summ <- rep(list(""), 2)
for (i in seq_along(negs_list)){
  if(length(negs_list[[i]])>0){
    xx <- rownames(tax_summ_matrix[[i]])[ceiling(negs_list[[i]]/nrow(tax_summ_matrix[[i]]))]
    figcapextra <- "As you can note, dashed lines and solid lines from trees coming out from both types of clustering algorithms implemented are mostly overlapping. This means that removing negative values does not change results from clustering algorithms much."
    cat(paste0("This taxon's ", names(negs_list)[i], " matrix has some negative values in the following taxa: *", paste(xx, collapse = "*, *"), "*. "))
  } else {
    figcapextra <- ""
    cat(paste("This taxon's",  names(negs_list)[i], "matrix has NO negative values."))
  }
  figcap_lttplot_summ[[i]] <- paste(taxon, "lineage through time (LTT) plots from source chronograms and",
  names(negs_list)[i], "summary matrix converted to phylo with different methods (NJ and UPGMA).",
  figcapextra, "Clustering algorithms used often are returning non-ultrametric trees or
  with maximum ages that are just off (too old or too young). So we developped an
  alternative algorithm in `datelife` to go from a summary matrix to a fully ultrametric tree.")
}
```
```{r LTTplot-median, echo = F, results = 'asis', fig.cap= ""}
figure <- paste0("\n![", figcap_lttplot_summ[[2]], "](plots/", taxon, "_LTTplot_Median.pdf)\n")
cat(figure)
```

```{r LTTplot-sdm, echo = F, results = 'asis', fig.cap= ""}
lttplot_sdm <- paste0("\n![", figcap_lttplot_summ[[1]], "](plots/", taxon, "_LTTplot_sdm.pdf)\n")
cat(lttplot_sdm)
```

```{r LTTplot-median-and-sdm-summtrees, echo = F, results = 'asis', fig.cap= ""}
figcap_lttplot_summ <- rep(list(""), 2)
for(i in 1:2){
  figcap_lttplot_summ[[i]] <- paste(taxon, "lineage through time (LTT) plots from source chronograms and",
  names(negs_list)[i], "summary matrix converted to phylo with `datelife` algorithm.")
}

lttplot_median <- paste0("\n![", figcap_lttplot_summ[[2]], "](plots/", taxon, "_LTTplot_summtrees_Median.pdf)\n")
lttplot_sdm <- paste0("\n![", figcap_lttplot_summ[[1]], "](plots/", taxon, "_LTTplot_summtrees_SDM.pdf)\n")
cat(lttplot_median)
cat(lttplot_sdm)
```
