---
title: "DateLife Workflows"
author: "Luna L. Sanchez Reyes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: library_red.bib
csl: systematic-biology.csl
link-citations: yes
header-includes:
- \usepackage{booktabs}
- \usepackage{makecell}
- \usepackage{multirow}
- \usepackage{longtable}
- \usepackage{caption}
- \usepackage{array}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{xcolor}
geometry: "left=2.5cm,right=2.5cm,top=2cm,bottom=2cm"
vignette: >
  %\VignetteIndexEntry{DateLife Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
# - \usepackage{tabu}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
my_table_format <- function(tt){
  res <- format.data.frame(tt, digits = 4, nsmall = 4, justify = "none")
  return(res)
}
library(microbenchmark)
loadd(tax_allcal)
loadd(taxon)
loadd(tax_dq)
loadd(tax_dr)
loadd(tax_summ)
loadd(tax_phyloall)
phyloall_success <- ape::is.ultrametric(tax_phyloall, option = 2)
loadd(tax_median_phylo)
loadd(tax_sdm_phylo)
loadd(tax_sdm_matrix)
loadd(tax_median_matrix)
loadd(tax_datedotol)
loadd(tax_allcal_datedotol)
loadd(tax_otol)
loadd(tax_treefromtax)
loadd(tax_phyloall)
loadd(tax_eachcal_datedotol)
loadd(tax_crossval)
loadd(tax_phycluster)
# loadd(tax_crossval2)
loadd(tax_summary)
loadd(tax_bestgrove)
fig_lttplot_phyloall <- "Fig. 1"
fig_lttplot_median <- "Fig. 2"
fig_lttplot_median_and_sdm <- "Fig. 3"
fig_node_age_dist <- "Fig. 4"
figcap2 <- paste(taxon, "Species Tree from Taxonomy. This tree was obtained with `tree_from_taxonomy()` function.")
table_chronograms <- "Table 1"
table_crossval <- "Table 2"
fig_lttplot_sdm <- "Fig. 5"
negs_list <- list(SDM = which(tax_sdm_matrix < 0),
                 Median = which(tax_median_matrix < 0))
tax_summ_matrix <- list(tax_sdm_matrix, tax_median_matrix)
negs_sdm <- c()
if(length(negs_list$SDM) > 0){
  xx <- rownames(tax_sdm_matrix)[ceiling(negs_list$SDM/nrow(tax_sdm_matrix))]
  negs_sdm <- paste0("In this case, the SDM summary matrix has some negative values in the following taxa: *", paste(xx, collapse = "*, *"), "*. ")
} else {
  negs_sdm <- paste0("But, the SDM summary matrix of this taxon has no negative values.")
}
```
\raggedright

# Taxon `r taxon`

## 1. Query source chronograms
There are `r length(tax_dq$cleaned_names)` species in the Open Tree of Life Taxonomy for the taxon `r taxon`.
Information on time of divergence is available for
```{r echo = FALSE, results = 'asis'}
tt <- nrow(tax_median_phylo$taxon_distribution)
if(tt == length(tax_dq$cleaned_names)){
  cat("all")
} else {
  cat(nrow(tax_median_phylo$taxon_distribution))
}
```
of these species across `r length(tax_dr)` published and peer-reviewed chronograms.
Original study citations as well as number of `r taxon` species found across those source
chronograms is shown in `r table_chronograms`.

```{r echo = FALSE, warning=FALSE, message = FALSE, results = 'asis'}
if(all(phyloall_success)){
  ultrametricity <- "All source chronograms are fully ultrametric"
} else {
  ultrametricity <- "Not all source chronograms are fully ultrametric. This cant't be right, CHECK THIS!!"
}
# Error producing PDF.
# ! Undefined control sequence.
# l.124 \toprule
# this occurs when the needed latex package is not specified in the yaml ^^
table_caption <- paste(taxon, "source chronogram studies information.")
make_table0(tax_dr, tax_summ, tax_dq, table_caption)
```

`r ultrametricity` and their maximum ages range from `r round(min(tax_summary$mrca), digits = 3)` to
`r round(max(tax_summary$mrca), digits = 3)` million years ago (MYA).
As a means for comparison, lineage through time plots of all source chronograms
available in data base are shown in `r fig_lttplot_phyloall`

\newpage

```{r LTTplot-phyloall, echo = F, results = 'asis', fig.cap= ""}
figcap_lttplot_phyloall <- paste0("Lineage through time (LTT) plots of source chronograms available in data base
  for species in the ", taxon, ". Numbers correspond to original studies in ", table_chronograms,
  ". Arrows indicate maximum age of each chronogram.")

lttplot_phyloall <- paste0("\n![", figcap_lttplot_phyloall, "](plots/", taxon, "_LTTplot_phyloall.pdf)\n")
cat(lttplot_phyloall)
```

## 2. Summarize results from query

LTT plots are a nice way to visually compare several trees. But what if you want
to summarize information from all source chronograms into a single summary chronogram?

The first step is to identify the degree of species overlap among your source chornograms: if each
source chronogram has a unique sample of species, it will not be possible to combine
them into a single summary chronogram. To identify the set of trees or _grove_ with the most source
chronograms that have at least two overlapping taxa, we followed Ané et al. 2016.
```{r echo = FALSE, warning=FALSE, message = FALSE, results = 'asis'}
if(length(tax_bestgrove) == length(tax_dr)){
    cat("In the case of the ", taxon, " all source chronograms have at least two overlapping species.\n")
    cat("\nNow that we know that the best grove has all source chronograms ")
} else {
    cat("In this case, not all source chronograms found for the ", taxon,
        " have at least two overlapping species. The largest grove has ",
            length(tax_bestgrove), " chronograms (out of ", length(tax_dr),
            " total source chronograms).\n")
    cat("\nNow that we have identified a grove ")
}
```
we can go on to summarize it by translating the source chronograms into patristic distance matrices and
then averaging them into a single summary matrix; yes, this first step is _that_
straightforward. We can average the source matrices by simply using the mean or
median distances, or we can use methods that involve transforming
the original distance matrices --such as the super distance matrix (SDM) approach of Criscuolo et al. 2006-- by minimizing
the distances across source matrices. As a result of such transformation, an SDM
summary matrix can contain negative values. `r negs_sdm`

Because our summary matrix is basically a distance matrix, a distance-based clustering
algorithm could be used to reconstruct the tree. Algorithms such as neighbour joining (NJ) and
unweighted pair group method with arithmetic mean (UPGMA) are
fast and work very well when there are no missing values in the matrices. However, summary
matrices coming from source chronograms usually have several NAs and missing rows.
<!--This data set for example has NUMBER cells with missing data.-->
When this happens, variants of traditional clustering algorithms have been developed to deal with missing values.
<!--
NJS: ape::njs
UPGMA: daisy + upgma
BIONJ: ape::bionjs(X)
Minimum Variance Reduction: ape::mvrs(X, V)
Triangle Method, Tree reconstruction based on the triangle method: ape::triangMtds(X)
-->
However, even these methods do not work well with our summary matrices, as shown in the following section.
We should note that these clustering methods are usually applied to distance matrices representing substitution rates
and not absolute time.

\newpage

###   2.1. Clustering a summary matrix

NJ, UPGMA, BIONJ, minimum variance reduction (MVR) and the triangle method (TM)
algorithms were used to cluster median and SDM summary distance matrices.
All clustering algorithms returned very similar trees with both types of summary
matrices (`r fig_lttplot_median`, Appendix `r fig_lttplot_sdm`).
UPGMA is the only algorithm that returns ultrametric trees, but they are considerably
older than expected from source chronograms.
The other methods returned trees with reasonable ages, but that are not ultrametric.

```{r make-lttplot-cluster-median, echo = F, results = 'asis', fig.cap= ""}
figcap_lttplot_cluster <- paste0("Lineage Through Time plots of ", taxon, " median summary
chronograms obtained with different clustering algorithms. Not all algorithms worked
with this summary matrix and we are only showing here the ones that worked.
Chronograms obtained from the SDM summary matrix are very similar to the ones from
the median summary matrix with all clustering algorithms (Appendix ", fig_lttplot_sdm, ").")

lttplot_cluster <- paste0("\n![", figcap_lttplot_cluster, "](plots/", taxon, "_lttplot_cluster_median.pdf)\n")
cat(lttplot_cluster)
```

An alternative to clustering algorithms is to use all data avilable in the summary
matrix as calibrations over a consensus tree.
The advantage of this is that we can get a distribution of ages for the nodes and
that we can essentially use this summary matrix to date any topology containing
at least some of the nodes, as shown in the `Create new data` section.

###   2.2. Calibrating a consensus tree with data from a summary matrix

Even if the branch lengths coming form the clustered chronograms are not adequate,
the topology can still be used as a consensus tree of the taxa with time data available.
Then, a list of divergence times available for each node can be constructed from the summary matrix,
simply by matching it to the node that corresponds to each pair of taxa in any given tree. Finally, the list
and consensus tree can be fed to any dating software that does not require data.
The branch length aduster (BLADJ) algorithm [@Webb2000] is really fast and does
not make any evolutionary assumptions on age distribution. Other software such as
MrBayes, r8s, and PATHd8 can be used instead of BLADJ by running them without data.
In here, we show summary chronograms obtained using minimum, mean and maximum distances
from summary matrices available for each node on the consensus tree and using them
as fixed ages in BLADJ (`r fig_lttplot_median_and_sdm`).
Chronograms from both types of summary matrices are quite similar. As expected,
SDM chronograms using minimum, mean and maximum distances do not vary much in their
maximum age, because ages are transformed to minimize variance across them. In contrast,
median chronogram obtained with minimum, mean and maximum distances have wider variation
in their maximum ages, as can be observed in the distance between the green arrows
in `r fig_lttplot_median_and_sdm`.

```{r LTTplot-summ-chronograms, echo = F, results = 'asis', fig.cap= ""}
figcap_lttplot_summchrono <- paste(taxon, "lineage through time (LTT) plots from
    source chronograms (gray), median (green) and SDM (blue) summary chronograms
    obtained by calibrating a consensus tree tropology with distance data
    from respective summary matrices and then adjusting branch lengths with BLADJ.")

lttplot_summchrono <- paste0("\n![", figcap_lttplot_summchrono, "](plots/", taxon, "_LTTplot_summary_chronograms.pdf)\n")
cat(lttplot_summchrono)
```


\newpage


## 3. Generate new chronograms

Another way to leverage information from the source chronograms is to use the node
ages as secondary calibration points to date any tree topology (with or without
branch lengths) given that at least two taxa from source chronograms are in
the tips of that topology.
In this data set, we have `r nrow(tax_allcal)` calibrations in total (that basically
corresponds to the sum of the number of nodes in each source chronogram).
Once we have a target tree topology, we can map the calibrations to the target tree.
Some nodes will have several calibrations and some others might have none.
To deal with this, we can expand the calibrations to make them agree, or we can summarize them.
(`r fig_node_age_dist`)

### 3.1. Expanding calibrations

show cross validation of LTTs from chronograms obtained by dating the topology of
each study with data from any other study.

### 3.2. Summarizing calibrations


### 3.3. Example with subspecies tree
As an example, we're gonna date the subspecies tree of the group (coming from otol).
```{r echo = FALSE}
txtA <- toupper("Something is missing.")
if(!inherits(tax_allcal_datedotol, "list")){
  txtA <- paste(taxon, "and no tree could be constructed.")
} else {
    if(is.null(tax_allcal_datedotol$phy)){
      txtA <- paste(taxon, "and they do not agree, so no tree could be constructed.")
    } else {
      if(all(is.na(tax_allcal_datedotol$phy$edge.length))){
        txtA <- paste(taxon, "and a tree was constructed, but all branch lengths are NA.")
      } else {
          txtA <- paste(taxon, "and a tree with", ape::Ntip(tax_allcal_datedotol$phy), "tips,",
                        round(tax_allcal_datedotol$phy$Nnode/ape::Ntip(tax_allcal_datedotol$phy)*100),
                        "% resolved nodes and a MRCA of",
                        round(max(ape::branching.times(tax_allcal_datedotol$phy))),"was constructed.")
      }
    }
}
```

Now, let's say you like the Open Tree of Life Taxonomy and you want to stick to
that tree. Dates from available studies were tested over the Open Tree of Life
Synthetic tree of `r txtA`
We also tried  each source chronogram independently, with the Dated OToL and with
each other, as a form of cross validation in `r table_crossval`. This is not working
perfectly yet, but we are developping new ways to use all calibrations efficiently.

\newpage
<!--
```{r echo = FALSE}
table_cap <- "Was it successful to use each source chronogram independently as calibration (CalibN) against the Dated Open Tree of Life (dOToL) and each other (ChronoN)?"
table2 <- cbind(tax_eachcal_datedotol, do.call(cbind, tax_crossval))
dimnames(table2) <- list(paste0("Calibrations", seq(tax_crossval)), c("dOToL", paste0("Chrono", seq(tax_crossval))))
if(ncol(table2) < 7) fs <- 10
if(ncol(table2) == 7 | ncol(table2) == 8) fs <- 9
if(ncol(table2) >= 9) fs <- 7
if(ncol(table2) > 10) {
  table2ori <- table2
  table2[table2 == TRUE] <- "yes" # "&#x2611;"  # :tick:
  table2[table2 == "FALSE"] <- "no"  # "&#x2612;"
  dimnames(table2) <- list(paste0("Calib", seq(tax_crossval)),
                           c("dOToL", paste0("Chr", seq(tax_crossval))))
}
tt <- knitr::kable(table2, caption = table_cap, row.names = TRUE, format = "latex", booktabs = TRUE)
kableExtra::kable_styling(kable_input = tt, position = "left", font_size = fs)
```
-->
\newpage
## 4. Simulate data/ Add missing taxa
An alternative to generate a dated tree from a set of taxa is to take the available information and simulate into it the missing data.
We will take the median and sdm summary chronograms to date the Synthetic tree of Life:
```{r LTTplot-sims, echo = F, results = 'asis', fig.cap= ""}
#nothing yet
```

\newpage
## References

## Appendix
```{r echo = FALSE, results = 'asis'}
absent <- levels(tax_median_phylo$absent_taxa$taxon)
if(!"None" %in% absent){
  cat("The following species were completely absent from the chronogram data base: ", paste0("*", levels(tax_median_phylo$absent_taxa$taxon), collapse = "*, *", "*"))
}
```

```{r make-lttplot-cluster-sdm, echo = F, results = 'asis', fig.cap= ""}
figcap_lttplot_cluster <- paste0("Lineage Through Time plots of ", taxon, " SDM summary
chronograms obtained with different clustering algorithms. Not all algorithms worked
with the SDM summary matrix and we are only showing here the ones that worked.
Chronograms obtained from the median summary matrix are very similar to the ones shown
here with all algorithms (main", fig_lttplot_median, ").")

lttplot_cluster <- paste0("\n![", figcap_lttplot_cluster, "](plots/", taxon, "_lttplot_cluster_median.pdf)\n")
cat(lttplot_cluster)
```


```{r myfile-1-plot, echo = F, results = 'asis', fig.cap= ""}
# knitr::include_graphics(paste0("docs/datedotol_", taxon, ".pdf"))
figcap <- paste(taxon, "Species Dated Open Tree of Life Induced Subtree. This chronogram was obtained with `get_dated_otol_induced_subtree()` function.")

if(!inherits(tax_datedotol, "phylo")){
  plot1 <- paste0("Dated induced subtree could not be obtained for the ", taxon, ".\n")
} else {
  plot1 <- paste0("\n![", figcap, "](plots/", taxon, "_datedotol.pdf)\n")
}
cat(plot1)
```


```{r eval = TRUE, echo = F, results = 'asis'}
figcap_lttplot_summ <- rep(list(""), 2)
for (i in seq_along(negs_list)){
  if(length(negs_list[[i]])>0){
    xx <- rownames(tax_summ_matrix[[i]])[ceiling(negs_list[[i]]/nrow(tax_summ_matrix[[i]]))]
    figcapextra <- "As you can note, dashed lines and solid lines from trees coming out from both types of clustering algorithms implemented are mostly overlapping. This means that removing negative values does not change results from clustering algorithms much."
    cat(paste0("This taxon's ", names(negs_list)[i], " matrix has some negative values in the following taxa: *", paste(xx, collapse = "*, *"), "*. "))
  } else {
    figcapextra <- ""
    cat(paste("This taxon's",  names(negs_list)[i], "matrix has NO negative values."))
  }
  figcap_lttplot_summ[[i]] <- paste(taxon, "lineage through time (LTT) plots from source chronograms and",
  names(negs_list)[i], "summary matrix converted to phylo with different methods (NJ and UPGMA).",
  figcapextra, "Clustering algorithms used often are returning non-ultrametric trees or
  with maximum ages that are just off (too old or too young). So we developped an
  alternative algorithm in `datelife` to go from a summary matrix to a fully ultrametric tree.")
}
```
```{r LTTplot-median, echo = F, results = 'asis', fig.cap= ""}
figure <- paste0("\n![", figcap_lttplot_summ[[2]], "](plots/", taxon, "_LTTplot_Median.pdf)\n")
cat(figure)
```

```{r LTTplot-sdm, echo = F, results = 'asis', fig.cap= ""}
lttplot_sdm <- paste0("\n![", figcap_lttplot_summ[[1]], "](plots/", taxon, "_LTTplot_sdm.pdf)\n")
cat(lttplot_sdm)
```

```{r LTTplot-median-and-sdm-summtrees, echo = F, results = 'asis', fig.cap= ""}
figcap_lttplot_summ <- rep(list(""), 2)
for(i in 1:2){
  figcap_lttplot_summ[[i]] <- paste(taxon, "lineage through time (LTT) plots from source chronograms and",
  names(negs_list)[i], "summary matrix converted to phylo with `datelife` algorithm.")
}

lttplot_median <- paste0("\n![", figcap_lttplot_summ[[2]], "](plots/", taxon, "_LTTplot_summtrees_Median.pdf)\n")
lttplot_sdm <- paste0("\n![", figcap_lttplot_summ[[1]], "](plots/", taxon, "_LTTplot_summtrees_SDM.pdf)\n")
cat(lttplot_median)
cat(lttplot_sdm)
```
