---
title: "DateLife Workflows"
author: "Luna L. Sanchez Reyes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
header-includes:
- \usepackage{booktabs}
- \usepackage{makecell}
- \usepackage{multirow}
- \usepackage{longtable}
- \usepackage{caption}
- \usepackage{array}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{xcolor}
geometry: "left=2.5cm,right=2.5cm,top=2cm,bottom=2cm"
vignette: >
  %\VignetteIndexEntry{DateLife Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
# - \usepackage{tabu}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
my_table_format <- function(tt){
  res <- format.data.frame(tt, digits = 4, nsmall = 4, justify = "none")
  return(res)
}
library(microbenchmark)
loadd(taxon)
loadd(tax_dq)
loadd(tax_dr)
loadd(tax_summ)
loadd(tax_phyloall)
phyloall_success <- ape::is.ultrametric(tax_phyloall, option = 2)
loadd(tax_median_phylo)
loadd(tax_sdm_phylo)
loadd(tax_sdm_matrix)
loadd(tax_datedotol)
loadd(tax_allcal_datedotol)
loadd(tax_otol)
loadd(tax_treefromtax)
loadd(tax_phyloall)
loadd(tax_eachcal_datedotol)
loadd(tax_crossval)
loadd(tax_phycluster)
# loadd(tax_crossval2)
loadd(tax_summary)
loadd(tax_bestgrove)
fig_lttplot_phyloall <- "Fig. 1"
fig_lttplot_median_and_sdm <- "Fig. 2"
fig_lttplot_median <- "Fig. 2"
figcap2 <- paste(taxon, "Species Tree from Taxonomy. This tree was obtained with `tree_from_taxonomy()` function.")
table_chronograms <- "Table 1"
table_crossval <- "Table 2"
fig_lttplot_sdm <- "Fig. S1"

```
\raggedright

# Taxon `r taxon`

## I. Query source data
There are `r length(tax_dq$cleaned_names)` species in the Open Tree of Life Taxonomy for the taxon `r taxon`.
Information on time of divergence is available for
```{r echo = FALSE, results = 'asis'}
tt <- nrow(tax_median_phylo$taxon_distribution)
if(tt == length(tax_dq$cleaned_names)){
  cat("all")
} else {
  cat(nrow(tax_median_phylo$taxon_distribution))
}
```
of these species across `r length(tax_dr)` published and peer-reviewed chronograms.
Original study citations as well as proportion of `r taxon` species found across those source
chronograms is shown in `r table_chronograms`.

```{r echo = FALSE, warning=FALSE, message = FALSE, results = 'asis'}
if(all(phyloall_success)){
  cat("All source chronograms are fully ultrametric.\n")
} else {
  cat("Not all source chronograms are fully ultrametric. This cant't be right, CHECK THIS!!\n")
}
# Error producing PDF.
# ! Undefined control sequence.
# l.124 \toprule
# this occurs when the needed latex package is not specified in the yaml ^^
table_caption <- paste(taxon, "source chronogram studies information.")
make_table0(tax_dr, tax_summ, tax_dq, table_caption)
```

Source chronograms maximum age range from `r round(min(tax_summary$mrca), digits = 3)` to
`r round(max(tax_summary$mrca), digits = 3)` million years ago (MYA).
As a means for comparison, lineage through time plots of all source chronograms
available in data base are shown in `r fig_lttplot_phyloall`

## II. Summarize results.

LTT plots are a nice way to visually compare several trees. But what if you want
to summarize information from all source chronograms into a single summary chronogram?

```{r LTTplot-phyloall, echo = F, results = 'asis', fig.cap= ""}
figcap_lttplot_phyloall <- paste0("Lineage through time (LTT) plots of source chronograms available in data base
  for species in the ", taxon, ". Numbers correspond to original studies in ", table_chronograms,
  ". Arrows indicate maximum age of each chronogram.")

lttplot_phyloall <- paste0("\n![", figcap_lttplot_phyloall, "](plots/", taxon, "_LTTplot_phyloall.pdf)\n")
cat(lttplot_phyloall)
```

The first step is to identify the degree of species overlap among your source chornograms: if each
source chronogram has a unique sample of species, it will not be possible to combine
them into a single summary chronogram. To identify the set of trees or _grove_ with the most source
chronograms that have at least two overlapping taxa, we followed Ané et al. 2016.
```{r echo = FALSE, warning=FALSE, message = FALSE, results = 'asis'}
if(length(tax_bestgrove) == length(tax_dr)){
    cat("In the case of the ", taxon, " all source chronograms have at least two overlapping species.\n")
    cat("Now that we know that the best grove has all source chronograms ")
} else {
    cat("In this case, not all source chronograms found for the ", taxon,
        " have at least two overlapping species. The largest grove has ",
            length(tax_bestgrove), " chronograms (out of ", length(tax_dr),
            " total source chronograms).\n")
    cat("Now that we have identified a suitable grove ")
}
```
we can go on to summarize it by translating the source chronograms into patristic distance matrices and
then averaging them into a single summary matrix; yes, this first step is _that_
straightforward. We can average the source matrices by simply using the mean or
median distances, or we can use methods that involve transforming
the original distance matrices --such as the super distance matrix (SDM) approach of Criscuolo et al. 2006-- by minimizing
the distances across source matrices.

Because our summary matrix is basically a distance matrix, a distance-based clustering
algorithm could be used to reconstruct the tree. Algorithms such as neighbour joining (NJ) and
unweighted pair group method with arithmetic mean (UPGMA) are
fast and work well when there are no missing values in the matrices. However, summary
matrices coming from source chronograms usually have several NAs and missing rows.
<!--This data set for example has NUMBER cells with missing data.-->
When this happens, clustering algorithms that have been developed to deal with missing values
<!--
NJS: ape::njs
UPGMA: daisy + upgma
BIONJ: ape::bionjs(X)
Minimum Variance Reduction: ape::mvrs(X, V)
Triangle Method, Tree reconstruction based on the triangle method: ape::triangMtds(X)
-->
do not work well, as shown in the following section. This is probbaly because these
methods are usually applied to distance matrices that represent evolutionary distance
in terms of sunstitution rate nad not absolute time, as is the case in here.


###   II.A. Detecting clustering issues.

We tested several clustering algorithms on summary distance matrices coming from median and SDM.
UPGMA returns ultrametric trees that are considerably older than source chronograms.
Even scaling the distance matrix down by a factor of 0.5 would not produce trees
with ages that are coherent with the source chronograms.
NJ non-ultrametric
(generally with NJ algorithms). In most studied cases, UPGMA returns
fully ultrametric trees but with very old ages (we had to multiply the matrix by
0.25 to get ages approximate to source chronograms ages, however this number is not justified,
it is just the number that approximates ages to source maximum ages the most). NJ returned reasonable
ages, but trees are way non ultrametric, as you can see in `r fig_lttplot_sdm`
and `r fig_lttplot_median`.

```{r make-lttplot-summ2-test, echo = F, results = 'asis', fig.cap= ""}
figcap_make_lttplot_summ2 <- paste("Lineage Through Time plots of", taxa, "summary
chronograms obtained from median (orange) and SDM (purple) summary matrices, clustered
using NJ (dashed lines) and UPGMA (solid line) algorithms.")

lttplot_make_lttplot_summ2 <- paste0("\n![", figcap_make_lttplot_summ2, "](plots/", taxon, "_make_lttplot_summ2_test.pdf)\n")
cat(lttplot_make_lttplot_summ2)
```

```{r eval = TRUE, echo = F, results = 'asis'}
negs_list <- list(SDM = which(tax_sdm_matrix < 0),
                 Median = which(tax_median_matrix < 0))
tax_summ_matrix <- list(tax_sdm_matrix, tax_median_matrix)
figcap_lttplot_summ <- rep(list(""), 2)
for (i in seq_along(negs_list)){
  if(length(negs_list[[i]])>0){
    xx <- rownames(tax_summ_matrix[[i]])[ceiling(negs_list[[i]]/nrow(tax_summ_matrix[[i]]))]
    figcapextra <- "As you can note, dashed lines and solid lines from trees coming out from both types of clustering algorithms implemented are mostly overlapping. This means that removing negative values does not change results from clustering algorithms much."
    cat(paste0("This taxon's ", names(negs_list)[i], " matrix has some negative values in the following taxa: *", paste(xx, collapse = "*, *"), "*. "))
  } else {
    figcapextra <- ""
    cat(paste("This taxon's",  names(negs_list)[i], "matrix has NO negative values."))
  }
  figcap_lttplot_summ[[i]] <- paste(taxon, "lineage through time (LTT) plots from source chronograms and",
  names(negs_list)[i], "summary matrix converted to phylo with different methods (NJ and UPGMA).",
  figcapextra, "Clustering algorithms used often are returning non-ultrametric trees or
  with maximum ages that are just off (too old or too young). So we developped an
  alternative algorithm in `datelife` to go from a summary matrix to a fully ultrametric tree.")
}
```

```{r LTTplot-median, echo = F, results = 'asis', fig.cap= ""}
figure <- paste0("\n![", figcap_lttplot_summ[[2]], "](plots/", taxon, "_LTTplot_Median.pdf)\n")
cat(figure)
```



###   II.B. Age distributions from Median and SDM summary trees.

Comparison of summary chronograms reconstructed with min and max ages.

```{r LTTplot-median-summtrees, echo = F, results = 'asis', fig.cap= ""}
for(i in 1:2){
  figcap_lttplot_summ[[i]] <- paste(taxon, "lineage through time (LTT) plots from source chronograms and", names(negs_list)[i], "summary matrix converted to phylo with `datelife` algorithm.")
}

lttplot_median <- paste0("\n![", figcap_lttplot_summ[[2]], "](plots/", taxon, "_LTTplot_summtrees_Median.pdf)\n")
cat(lttplot_median)
```


```{r LTTplot-sdmsummtrees, echo = F, results = 'asis', fig.cap= ""}
lttplot_sdm <- paste0("\n![", figcap_lttplot_summ[[1]], "](plots/", taxon, "_LTTplot_summtrees_SDM.pdf)\n")
cat(lttplot_sdm)
```


\newpage


## III. Create new data


As an example, we're gonna date the Open Tree Synthetic tree (mainly because the taxonomic tree is usually less well resolved.)
```{r echo = FALSE}
txtA <- toupper("Something is missing.")
if(!inherits(tax_allcal_datedotol, "list")){
  txtA <- paste(taxon, "and no tree could be constructed.")
} else {
    if(is.null(tax_allcal_datedotol$phy)){
      txtA <- paste(taxon, "and they do not agree, so no tree could be constructed.")
    } else {
      if(all(is.na(tax_allcal_datedotol$phy$edge.length))){
        txtA <- paste(taxon, "and a tree was constructed, but all branch lengths are NA.")
      } else {
          txtA <- paste(taxon, "and a tree with", ape::Ntip(tax_allcal_datedotol$phy), "tips,",
                        round(tax_allcal_datedotol$phy$Nnode/ape::Ntip(tax_allcal_datedotol$phy)*100),
                        "% resolved nodes and a MRCA of",
                        round(max(ape::branching.times(tax_allcal_datedotol$phy))),"was constructed.")
      }
    }
}
```

Now, let's say you like the Open Tree of Life Taxonomy and you want to stick to that tree. Dates from available studies were tested over the Open Tree of Life Synthetic tree of `r txtA`
We also tried  each source chronogram independently, with the Dated OToL and with each other, as a form of cross validation in `r table_crossval`. This is not working perfectly yet, but we are developping new ways to use all calibrations efficiently.

\newpage
```{r echo = FALSE}
table_cap <- "Was it successful to use each source chronogram independently as calibration (CalibN) against the Dated Open Tree of Life (dOToL) and each other (ChronoN)?"
table2 <- cbind(tax_eachcal_datedotol, do.call(cbind, tax_crossval))
dimnames(table2) <- list(paste0("Calibrations", seq(tax_crossval)), c("dOToL", paste0("Chrono", seq(tax_crossval))))
if(ncol(table2) < 7) fs <- 10
if(ncol(table2) == 7 | ncol(table2) == 8) fs <- 9
if(ncol(table2) >= 9) fs <- 7
if(ncol(table2) > 10) {
  table2ori <- table2
  table2[table2 == TRUE] <- "yes" # "&#x2611;"  # :tick:
  table2[table2 == "FALSE"] <- "no"  # "&#x2612;"
  dimnames(table2) <- list(paste0("Calib", seq(tax_crossval)),
                           c("dOToL", paste0("Chr", seq(tax_crossval))))
}
tt <- knitr::kable(table2, caption = table_cap, row.names = TRUE, format = "latex", booktabs = TRUE)
kableExtra::kable_styling(kable_input = tt, position = "left", font_size = fs)
```
\newpage
## III. Simulate data
An alternative to generate a dated tree from a set of taxa is to take the available information and simulate into it the missing data.
We will take the median and sdm summary chronograms to date the Synthetic tree of Life:
```{r LTTplot-sims, echo = F, results = 'asis', fig.cap= ""}
#nothing yet
```

\newpage

## Appendix
```{r echo = FALSE, results = 'asis'}
absent <- levels(tax_median_phylo$absent_taxa$taxon)
if(!"None" %in% absent){
  cat("The following species were completely absent from the chronogram data base: ", paste0("*", levels(tax_median_phylo$absent_taxa$taxon), collapse = "*, *", "*"))
}
```
```{r myfile-1-plot, echo = F, results = 'asis', fig.cap= ""}
# knitr::include_graphics(paste0("docs/datedotol_", taxon, ".pdf"))
figcap <- paste(taxon, "Species Dated Open Tree of Life Induced Subtree. This chronogram was obtained with `get_dated_otol_induced_subtree()` function.")

if(!inherits(tax_datedotol, "phylo")){
  plot1 <- paste0("Dated induced subtree could not be obtained for the ", taxon, ".\n")
} else {
  plot1 <- paste0("\n![", figcap, "](plots/", taxon, "_datedotol.pdf)\n")
}
cat(plot1)
```


```{r LTTplot-sdm, echo = F, results = 'asis', fig.cap= ""}
lttplot_sdm <- paste0("\n![", figcap_lttplot_summ[[1]], "](plots/", taxon, "_LTTplot_sdm.pdf)\n")
cat(lttplot_sdm)
```
